Speeding up eQTL scans in the BXD population using GPU

### 2019-09-30 paper.tex
added cuarray reference, check off some todo

@@

\begin{abstract}
The BXD strains of mice are an important reference population for
systems biology and genetics that have been full sequenced and deeply phenotyped.
To facilitate interactive use of genotype-phenotype relationships
in this family, we sought to develop new algorithms and code to enable
high performance whole genome QTL scan for massive omics data sets of up to 1 million 
traits. By using easily parallelizable operations including
matrix multiplication, vectorized operations, and elementwise
operations, we have been able to decrease runtimes approaching real-time
computation, making this code ideal for interactive web services, such as
GeneNetwork.org. We used parallelization of different CPU threads
as well as GPUs. We found that the speed advantage of GPUs is
dependent on problem size and shape (n of cases, n of genotypes, n of traits).  
Our results provide a clear path for speeding up eQTL scans using LMMs. 
Our implementation is in the Julia programming language.
\end{abstract}

@@

We sought to build upon these efforts aiming to perform real-time eQTL
scans for the BXD populations.  We used the above-mentioned ideas and
examined the use of graphical processing units (GPUs) to speed these
further.  Since programming for GPUs is often non-trivial needing the
use of low-level languages such as C++, we used the Julia programming
language which offers a GPU programming capabilities while retaining
the simplicity of a high-level language such as R or Matlab.  Finally,
since most phenotype-marker associations are null, we examined the
impact of storage precision, and only returning the highest
association (LOD) score for each phenotype instead of a matrix of LOD
scores for every pair of marker and phenotypes.  With these efforts,
we find that it is possible to perform perform real-time eQTL scans
for the BXD population.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natural to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
They target different hardware or use various techniques to get highly optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded and therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays affords much lower programming efforts compared to C/C++.
Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

\inputminted{julia}{code/gpu.jl}
%TODO: Add reference to CUDAnative and CuArrays

@@

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standardized matrices ($G^{\ast}$, $Y^{\ast}$) for input matrices ($G$, $Y$)
	\item Get a correlation matrix ($R$) by multiplying the standardized matrices
	\item Calculate LOD scores 
\end{itemize}
Our profiling result shows that the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of the performance time.  That is, the former is rather slower than the latter regardless of precision for the data type.
We used a GPU profiler \textit{nvprof} \cite{nvprof}
to investigate what is the bottleneck of GPU. 
The result is, 98\% of the time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, the Y' and G matrix is about 40MB %TODO: double check on this number 
while R matrix is about 4GB. 
Data offloading is the main bottleneck for our GPU implementation.

>>>>

% !TeX document-id = {a69d50eb-b7fd-4cc1-86b3-4b24edd99f0d}
% !TeX TXS-program:compile = txs:///pdflatex/[--shell-escape]

@@

\begin{abstract}
The BXD recombinant inbred strains of mice are an important reference population for
systems biology and genetics that have been full sequenced and deeply phenotyped.
To facilitate interactive use of genotype-phenotype relationships
in this family, we sought to develop new algorithms and code to enable
high performance whole genome QTL scan for massive omics data sets of up to 1 million 
traits. By using easily parallelizable operations including
matrix multiplication, vectorized operations, and elementwise
operations, we have been able to decrease runtimes approaching real-time
computation, making this code ideal for interactive web services, such as
GeneNetwork.org. We used parallelization of different CPU threads
as well as GPUs. We found that the speed advantage of GPUs is
dependent on problem size and shape (n of cases, n of genotypes, n of traits).  
Our results provide a clear path for speeding up eQTL scans using Linear Mixed Model. 
Our implementation is in the Julia programming language.
\end{abstract}

@@

We sought to build upon these efforts aiming to perform real-time eQTL
scans for the BXD populations.  We used the above-mentioned ideas and
examined the use of graphical processing units (GPUs) to speed these
further.  Since programming for GPUs is often non-trivial needing the
use of low-level languages such as C++, we used the Julia programming
language which offers a GPU programming capabilities while retaining
the simplicity of a high-level language such as R or Matlab.  Finally,
since most phenotype-marker associations are null, we examined the
impact of storage precision, and only returning the highest
association (LOD) score for each phenotype instead of a matrix of LOD
scores for every pair of marker and phenotypes.  With these efforts,
we find that it is possible to perform real-time eQTL scans
for the BXD population.

@@

% OpenBLAS for multi-threaded matrix operations. 
Since our algorithm largely depends on matrix operations, it is natural to find the fastest way to achieve best result, regardless of computing platform. 
There are various matrix libraries available for CPU, such as gslBLAS and OpenBLAS \citep{wang2013augem}. 
They target different hardware or use various techniques to get optimal results. 
The matrix multiplication in OpenBLAS is default multithreaded and therefore does not require extra coding effort to make the CPU version of matrix multiplication run in parallel.
We chose OpenBLAS as CPU computing library.

@@

\subsubsection{Julia language}
%Programming simplicity of interpreted languages with speed approaching compiled languages; GPU programming and linear algebra straightforward
In our experiment of matrix multiplication, Julia's speed is comparable to C/C++. 
However, the low learning curve, clean syntax, as well as support to GPU programming libraries such as CUDAnative \citep{CUDAnative.jl-2018} and CuArrays \citep{cuarray} affords much lower programming efforts compared to C/C++.
Compared to writing GPU functions in C/C++, writing in Julia is cleaner and easier because it requires much less boiler plate code.
Below is some code snippets to demonstrate GPU code, calling library or writing custom kernel in Julia.

\inputminted{julia}{code/gpu.jl}

@@

The genome scan process includes the following steps:
\begin{itemize}
	\item Calculate standardized matrices ($G^{\ast}$, $Y^{\ast}$) for input matrices ($G$, $Y$)
	\item Get a correlation matrix ($R$) by multiplying the standardized matrices
	\item Calculate LOD scores 
\end{itemize}
Our profiling result shows that the second and third steps take up 90\% of the time and involve parallelizable matrix operations.  Hence, they are our candidates for GPU acceleration.
To make a fair comparison of the CPU and GPU, the timing shown in Table \ref{tab:speedup-table} is the total execution time for genome scan. 
The \textit{No Find Max LOD} method shows the timing when running the scan with \textit{CPU only} and \textit{CPU\&GPU}.
As the table shows, the CPU\&GPU combination did not show any improvement over the serial version (CPU only) in terms of the performance time.  That is, the former is rather slower than the latter regardless of precision for the data type.
We used a GPU profiler \textit{nvprof} \cite{nvprof}
to investigate what is the bottleneck of GPU. 
The result is, 98\% of the time is spent on data transfer from GPU to CPU (device to host).
As shown in Figure \ref{MatrixMult}, the input matrices Y' and G are small compared to the output matrix R.
For BXD spleen dataset, Y' matrix is 17 MB, G matrix is 21 MB
while R matrix is about 4GB. 
Data offloading is the main bottleneck for our GPU implementation.
